Multi-redshift limits on the 21cm power spectrum from PAPER

### 2014-10-21 ms.tex
removing comments for post to arxiv

\documentclass[fleqn,10pt,iop]{emulateapj}
%\documentclass[preprint2]{aastex}

@@

%
%This signal is though to be the richest data set on the sky!  Many telescopes are searching for the power spectral signature of 
%HI.  Coarsely, models of 21cm emission can be distilled into two parameters. The redshift at which point the universe was 50\% ionized ($z_i$) and the transition time from mostly neutral to mostly ionized $dz$. Thus full coverage of the spectrum is essential.

%This paper extends the result ARP2013a to cover the redshift range XXX.

@@

\subsection{Initial Averaging}
  First, the raw data are down-selected to just the 70 $\sim$30m long baselines described in Section \ref{sec:observations}.  %I know this isn't technically true, but I think its important that the downselect is obvious up front.
   The visibilities are then compressed in the frequency and time directions by removing delay modes and highest frequency fringe rates corresponding to a 300m baseline (the longest baseline in the array).  This filtering\footnote{See Appendix A of P14} is done in tandem with a radio frequency interference (RFI) flagging step, using the residuals which have had bright sky-like signals removed to flag 4$\sigma$ deviations before feeding the flags back into another iteration of the compression step. This mitigates the effects of bright, narrow band, interference  being scattered into higher delay modes (where only the reionization signal is expected to be found) and results in a time and frequency bin of length 40s and width 492.61kHz. 
  This process reduces the data volume by a factor of $\sim$20, or roughly an order of magnitude improvement on traditional time and frequency averaging which in this array would be limited to ~100kHz and 10s to avoid averaging away fringes.

@@

 \begin{figure*}
\centering
\includegraphics[width=\textwidth]{figures/psa32_flagging_zbins.png}
\caption{\label{fig:bins} The average amount of data remaining after interference flagging over the 3 month period between Dec 2011 and March 2012 (black line) is quite high.  Redshift bins (in grey, redshift center indicated with text label) are chosen to include spectral channels with uniform weight, i.e. no missing channels while maximizing coverage over the band.  Redshift 7.68 is included for comparison with P14.  Channels with no missing data are indicated by the dotted line, visible at the edge of flagged channels. Each redshift bin is 20 MHz wide, but weighted by a Blackman-Harris window function which heavily down-weights the outer 10MHz for a Noise Equivalent Bandwidth of 10MHz.  The interference is almost exclusively dominated by two features: ORBCOMM satellites at 137MHz and an unidentified intermittent line emitter at 175MHz. The roll off at 115MHz is due the rising noise at low frequencies being incorrectly flagged as interference.  }
\end{figure*} 
%The success of the remainder of the processing stages hinges on the spectral stability of the data, which can easily be assessed by comparing redundant baselines. In figure XXX we see the r.m.s. of the ratios between redundant calibrations, before and after the calibration has been applied.  
\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/psa32_trms_with_and_without_fr_filtering_z.png}
\caption{\label{fig:noise} Root mean square (rms) noise before and after integrating to 13 minute LST bins (top and bottom sets of curves) indicating the rising significance of foregrounds as we take the last deep integration step. Noise is estimated by differencing adjacent frequencies (magenta) and between redundant baselines (blue), compared with an estimate of the theoretical noise level with a $T_{sys}$ of 550K (dotted). The top three lines show noise after filtering foregrounds and binning into 40s long sidereal bins.  At this noise level the frequency and baseline differences are roughly similar, both demonstrate the same small bumps of increased noise due to interference flagging and are consistent with noise over much of the band.  The bottom three lines show the noise level after integrating up to the maximum fringe rate of 789s. The step between the top and bottom sets is the last coherent integration with noise largely decreasing by the expected factor $\sqrt{789/40}$ except in the difference between baselines (blue) which demonstrates a clear excess at all frequencies, particularly above redshift 10. With this last large jump in sensitivity we are now seeing the slight dominance of baseline covariance over the spectral derivative rms. }
\end{figure}

@@

%This step is the first point at which signal loss could occur. To measure the loss we add white noise into the data at a level sufficient to double the system temperature (to make it easily detectable on the output), subtract  and then measure the fractional difference post filtration. We recover XXX\% of the injected signal.

%delay transform, covariance projection (lossless, and lossy), bin and bootstrap average,
\subsection{Power Spectrum}
\label{sec:power_spectrum}
The output of the above steps is eight sidereal hours of calibrated, foreground filtered visibility data averaged over 92 nights. The power spectrum is estimated in the delay spectrum of  a 10MHz bandwidth range centered on the redshift of interest. To preserve the isolation of any foregrounds which remain, we increase the spectral range by 5MHz on each side and multiply by a Blackman-Harris window thus providing a much higher dynamic range delay spectrum point spread function.

@@

The residual elements of the correlation matrix corresponding to cross-multiplication of matching delay bins between different baselines are all un-biased samples of the power spectrum. To estimate the final power spectrum and its uncertainty we compute the mean and variance of many random randomly-selected subsamples, sampling the dimensions: sidereal time, redundant baseline pair, and delay sign\footnote{As visibilities are complex, both the positive and negative delays  carry  information. Physically the two signs correspond to the two halves of the sky.}.

@@

The theoretical noise power spectrum (dashed line in Figures \ref{fig:pspecs} and \ref{fig:P14compare} and grey region in Figure \ref{fig:slices}) is estimated using the method described in \cite{Pober:2013p9581}, assuming a system temperature of 550K and the observing scheme described in Section \ref{sec:observations}. In all figures the noise levels are plotted for comparison with the power spectrum values rather than the error bars, i.e. 68\% of uncorrelated, noise-dominated points should lie below the line or within the grey region.    Points where the error bar does not cross zero are inconsistent with noise at $>68$\% these are listed as ``detections'' in Table \ref{tab:data}.

If our measurements were completely noise dominated, $68\%$ of the data points ought to include zero within their error bars.  In Table \ref{tab:data}, we divide the points into ``Detections" (power spectrum values smaller than $1\sigma$ errors), and ``Upper limits" (measured values larger).  With only one exception, this scheme yields the same results whether we use the theoretical errors or the bootstrap errors.  Under this scheme, we would expect a noise dominated signal to yield around eight ``upper limits'' with an equal number of positive and negative data points. Here we find only two to three upper limits while rest we classify as detections. In general, most detections are within 1 - 2 $\sigma$ but are all positively biased,  with the significance rising steeply at high redshifts and low $k$s.%, the quadratic sum of all errors above $k=0.1$ is 6.6 $\sigma$ indicating that the excess is largely coherent across the data set.

\subsection{The Excess Signal}

@@

\begin{deluxetable}{llccccccc}
\tablecolumns{9}
\tablecaption{PAPER Power spectrum values using 32 antenna and a full season of integration.}
\tablehead{
\colhead{k} & 
\colhead{redshift}  & 
\colhead{$\Delta^2$} &
\colhead{\parbox{6em}{theoretical error} }&
\colhead{\parbox{6em}{theoretical significance}} &
\colhead{\parbox{6em}{theoretical detection}} &
\colhead{\parbox{6em}{bootstrap error }} &
\colhead{\parbox{6em}{bootstrap significance}} &
 \colhead{\parbox{5em}{bootstrap detection}} \\
\colhead{[hMpc$^{-1}$]} & 
\colhead{}  & 
\colhead{[mK$^2$]} &
\colhead{ $\sigma$} & 
\colhead{}&
\colhead{\parbox{7em}{[mK$^2$]} }& 
\colhead{[mK$^2$]}&
\colhead{$\sigma$} & 
\colhead{}
}
\startdata
0.1 & 10.29 & 43600 & $\pm$   700 & 64.4 & Det & $\pm$  5700 & 7.6 & Det \tabularnewline
0.1 & 8.54 &  5700 & $\pm$   300 & 20.4 & Det & $\pm$  1200 & 4.9 & Det \tabularnewline
0.1 & 7.94 &  7700 & $\pm$   200 & 33.6 & Det & $\pm$  2000 & 3.8 & Det \tabularnewline
0.1 & 7.55 &  9500 & $\pm$   200 & 61.7 & Det & $\pm$  2800 & 3.4 & Det \tabularnewline
0.2 & 10.29 & 11400 & $\pm$  5300 & 2.1 & Det & $\pm$  5000 & 2.3 & Det \tabularnewline
0.2 & 8.54 &  5900 & $\pm$  2200 & 2.7 & Det & $\pm$  2200 & 2.7 & Det \tabularnewline
0.2 & 7.94 &  5100 & $\pm$  1800 & 2.9 & Det & $\pm$  1800 & 2.9 & Det \tabularnewline
0.2 & 7.55 &   900 & $\pm$  1200 & 0.7 & ULim & $\pm$  1500 & 0.6 & ULim \tabularnewline
0.3 & 10.29 & 34800 & $\pm$ 17900 & 1.9 & Det & $\pm$ 16400 & 2.1 & Det \tabularnewline
0.3 & 8.54 & 12900 & $\pm$  7300 & 1.8 & Det & $\pm$  5400 & 2.4 & Det \tabularnewline
0.3 & 7.94 &  5200 & $\pm$  6000 & 0.9 & ULim & $\pm$  4800 & 1.1 & Det \tabularnewline
0.3 & 7.55 &  7600 & $\pm$  4000 & 1.9 & Det & $\pm$  4800 & 1.6 & Det \tabularnewline
0.4 & 10.29 & 42900 & $\pm$ 42400 & 1.0 & Det & $\pm$ 30700 & 1.4 & Det \tabularnewline
0.4 & 8.54 & 24100 & $\pm$ 17400 & 1.4 & Det & $\pm$ 16600 & 1.5 & Det \tabularnewline
0.4 & 7.94 & 14700 & $\pm$ 14200 & 1.0 & Det & $\pm$  9900 & 1.5 & Det \tabularnewline
0.4 & 7.55 & -8600 & $\pm$  9400 & -0.9 & ULim & $\pm$ 14200 & -0.6 & ULim \tabularnewline
\enddata
\tablenotetext{a}{All error bars are 1 $\sigma$.}
\tablenotetext{b}{Det indicates a measurement and error-bar inconsistent with zero, ULim indicates consistency with zero at 1 $\sigma$.}
\label{tab:data}
\end{deluxetable}

@@

Further work will look to expand on this result on several fronts.  Alternative algorithms to the wide-band CLEAN could potentially better remove the delay-space covariance introduced by RFI flags in frequency and are under investigation.  Another active point of investigation is aimed at alternative windows to the Blackman-Harris used in the wide-band CLEAN (or wholly distinct algorithms) that could better remove foregrounds from the edges of the band while still limiting foreground bleed into the EoR window.

>>>>

%\documentclass[fleqn,10pt,iop]{emulateapj}
\documentclass[preprint2]{aastex}

@@

@@

\subsection{Initial Averaging}
  First, the raw data are down-selected to just the 70 $\sim$30m long baselines described in Section \ref{sec:observations}.  
   The visibilities are then compressed in the frequency and time directions by removing delay modes and highest frequency fringe rates corresponding to a 300m baseline (the longest baseline in the array).  This filtering\footnote{See Appendix A of P14} is done in tandem with a radio frequency interference (RFI) flagging step, using the residuals which have had bright sky-like signals removed to flag 4$\sigma$ deviations before feeding the flags back into another iteration of the compression step. This mitigates the effects of bright, narrow band, interference  being scattered into higher delay modes (where only the reionization signal is expected to be found) and results in a time and frequency bin of length 40s and width 492.61kHz. 
  This process reduces the data volume by a factor of $\sim$20, or roughly an order of magnitude improvement on traditional time and frequency averaging which in this array would be limited to ~100kHz and 10s to avoid averaging away fringes.

@@

 \begin{figure*}
\centering
\includegraphics[width=\textwidth]{figures/psa32_flagging_zbins.png}
\caption{\label{fig:bins} The average amount of data remaining after interference flagging over the 3 month period between Dec 2011 and March 2012 (black line) is quite high.  Redshift bins (in grey, redshift center indicated with text label) are chosen to include spectral channels with uniform weight, i.e. no missing channels while maximizing coverage over the band.  Redshift 7.68 is included for comparison with P14.  Channels with no missing data are indicated by the dotted line, visible at the edge of flagged channels. Each redshift bin is 20 MHz wide, but weighted by a Blackman-Harris window function which heavily down-weights the outer 10MHz for a Noise Equivalent Bandwidth of 10MHz.  The interference is almost exclusively dominated by two features: ORBCOMM satellites at 137MHz and an unidentified intermittent line emitter at 175MHz. The roll off at 115MHz is due the rising noise at low frequencies being incorrectly flagged as interference.  }
\end{figure*}

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/psa32_trms_with_and_without_fr_filtering_z.png}
\caption{\label{fig:noise} Root mean square (rms) noise before and after integrating to 13 minute LST bins (top and bottom sets of curves) indicating the rising significance of foregrounds as we take the last deep integration step. Noise is estimated by differencing adjacent frequencies (magenta) and between redundant baselines (blue), compared with an estimate of the theoretical noise level with a $T_{sys}$ of 550K (dotted). The top three lines show noise after filtering foregrounds and binning into 40s long sidereal bins.  At this noise level the frequency and baseline differences are roughly similar, both demonstrate the same small bumps of increased noise due to interference flagging and are consistent with noise over much of the band.  The bottom three lines show the noise level after integrating up to the maximum fringe rate of 789s. The step between the top and bottom sets is the last coherent integration with noise largely decreasing by the expected factor $\sqrt{789/40}$ except in the difference between baselines (blue) which demonstrates a clear excess at all frequencies, particularly above redshift 10. With this last large jump in sensitivity we are now seeing the slight dominance of baseline covariance over the spectral derivative rms. }
\end{figure}

@@

\subsection{Power Spectrum}
\label{sec:power_spectrum}
The output of the above steps is eight sidereal hours of calibrated, foreground filtered visibility data averaged over 92 nights. The power spectrum is estimated in the delay spectrum of  a 10MHz bandwidth range centered on the redshift of interest. To preserve the isolation of any foregrounds which remain, we increase the spectral range by 5MHz on each side and multiply by a Blackman-Harris window thus providing a much higher dynamic range delay spectrum point spread function.

@@

The residual elements of the correlation matrix corresponding to cross-multiplication of matching delay bins between different baselines are all un-biased samples of the power spectrum. To estimate the final power spectrum and its uncertainty we compute the mean and variance of many random randomly-selected subsamples, sampling the dimensions: sidereal time, redundant baseline pair, and delay sign\footnote{As visibilities are complex, both the positive and negative delays  carry  information. Physically the two signs correspond to the two halves of the sky.}. For simplicity, measurements are averaged with equal weights.

@@

The theoretical noise power spectrum (dashed line in Figures \ref{fig:pspecs} and \ref{fig:P14compare} and grey region in Figure \ref{fig:slices}) is estimated using the method described in \cite{Pober:2013p9581}, assuming a system temperature of 550K and the observing scheme described in Section \ref{sec:observations}. In general the theoretical error bars are smaller than those derived by the bootstrap.  This is most likely due to the use of uniform weighting when combining power spectrum measurements, rather then the inverse variance weights assumed in the sensitivity calculation.

 In all figures the noise levels are plotted for comparison with the power spectrum values rather than the error bars, i.e. 68\% of uncorrelated, noise-dominated points should lie below the line or within the grey region.    Points where the error bar does not cross zero are inconsistent with noise at $>68$\% these are listed as ``detections'' in Table \ref{tab:data}.

If our measurements were completely noise dominated, $68\%$ of the data points ought to include zero within their error bars.  In Table \ref{tab:data}, we divide the points into ``Detections" (power spectrum values smaller than $1\sigma$ errors), and ``Upper limits" (measured values larger).  With only one exception, this scheme yields the same results whether we use the theoretical errors or the bootstrap errors.  Under this scheme, we would expect a noise dominated signal to yield around eight ``upper limits'' with an equal number of positive and negative data points. Here we find only two upper limits while the rest we classify as low significance detections of excess power. In general, most detections are within 1 - 2 $\sigma$ but are all positively biased,  with the significance rising steeply at high redshifts and low $k$s.%, the quadratic sum of all errors above $k=0.1$ is 6.6 $\sigma$ indicating that the excess is largely coherent across the data set.

\subsection{The Excess Power}

@@

\begin{deluxetable}{llccccccc}
\tablecolumns{9}
\tablecaption{PAPER Power spectrum values using 32 antenna and a full season of integration.}
\tablehead{
\colhead{k} & 
\colhead{redshift}  & 
\colhead{$\Delta^2$} &
\colhead{\parbox{6em}{theoretical error} }&
\colhead{\parbox{6em}{theoretical significance}} &
\colhead{\parbox{6em}{theoretical detection}} &
\colhead{\parbox{6em}{bootstrap error }} &
\colhead{\parbox{6em}{bootstrap significance}} &
 \colhead{\parbox{5em}{bootstrap detection}} \\
\colhead{[hMpc$^{-1}$]} & 
\colhead{}  & 
\colhead{[mK$^2$]} &
\colhead{[mK$^2$]} &
\colhead{ $\sigma$} & 
\colhead{}&
\colhead{\parbox{7em}{[mK$^2$]} }& 
\colhead{$\sigma$} & 
\colhead{}
}
\startdata
0.1 & 10.29 & 43600 & $\pm$   400 & 105.6 & Det & $\pm$  5700 & 7.6 & Det \tabularnewline
0.1 & 8.54 &  5700 & $\pm$   200 & 33.0 & Det & $\pm$  1200 & 4.9 & Det \tabularnewline
0.1 & 7.94 &  7700 & $\pm$   100 & 56.2 & Det & $\pm$  2000 & 3.8 & Det \tabularnewline
0.1 & 7.55 &  9500 & $\pm$   100 & 122.7 & Det & $\pm$  2800 & 3.4 & Det \tabularnewline
0.2 & 10.29 & 11400 & $\pm$  3300 & 3.5 & Det & $\pm$  5000 & 2.3 & Det \tabularnewline
0.2 & 8.54 &  5900 & $\pm$  1400 & 4.3 & Det & $\pm$  2200 & 2.7 & Det \tabularnewline
0.2 & 7.94 &  5100 & $\pm$  1100 & 4.8 & Det & $\pm$  1800 & 2.9 & Det \tabularnewline
0.2 & 7.55 &   900 & $\pm$   600 & 1.4 & Det & $\pm$  1500 & 0.6 & ULim \tabularnewline
0.3 & 10.29 & 34800 & $\pm$ 11000 & 3.2 & Det & $\pm$ 16400 & 2.1 & Det \tabularnewline
0.3 & 8.54 & 12900 & $\pm$  4600 & 2.8 & Det & $\pm$  5400 & 2.4 & Det \tabularnewline
0.3 & 7.94 &  5200 & $\pm$  3600 & 1.4 & Det & $\pm$  4800 & 1.1 & Det \tabularnewline
0.3 & 7.55 &  7600 & $\pm$  2000 & 3.8 & Det & $\pm$  4800 & 1.6 & Det \tabularnewline
0.4 & 10.29 & 42900 & $\pm$ 26000 & 1.7 & Det & $\pm$ 30700 & 1.4 & Det \tabularnewline
0.4 & 8.54 & 24100 & $\pm$ 10800 & 2.2 & Det & $\pm$ 16600 & 1.5 & Det \tabularnewline
0.4 & 7.94 & 14700 & $\pm$  8500 & 1.7 & Det & $\pm$  9900 & 1.5 & Det \tabularnewline
0.4 & 7.55 & -8600 & $\pm$  4800 & -1.8 & Det & $\pm$ 14200 & -0.6 & ULim \tabularnewline
\enddata
\tablenotetext{a}{All error bars are 1 $\sigma$.}
\tablenotetext{b}{Det indicates a measurement and error-bar inconsistent with zero, ULim indicates consistency with zero at 1 $\sigma$.}
\label{tab:data}
\end{deluxetable}

@@

Further work will look to expand on this result on several fronts.  Alternative algorithms to the wide-band CLEAN could potentially better remove the delay-space covariance introduced by RFI flags in frequency and are under investigation.  Another active point of investigation is aimed at alternative windows to the Blackman-Harris used in the wide-band CLEAN (or wholly distinct algorithms) that could better remove foregrounds from the edges of the band while still limiting foreground bleed into the EoR window.  Weighting of the final power spectrum average is also a subject for further improvement.   In this analysis we have used uniform weights; this is also the default assumption of most sensitivity calculations, including the one used here.  Uniform weighting is optimal if the final measurements are limited by something other than noise, while inverse variance is appropriate for purely noise limited measurements.  Given how close the error bars are to noise dominated, future analysis will investigate using different estimates of the variance for optimal weighting.
